OpenMPIR

### 2017-09-05 paper.tex
Updated latest figures, added some discussion

The remainder of the paper is organized as follows: In Section~\ref{Sec:background} 
we give background and motivation for the paper. Following that, we describe OpenMP
tasking in Section~\ref{Sec:OpenMP}, and Tapir in Section~\ref{Sec:Tapir}. We then 
discuss the implementation, and how we compile OpenMP task constructs to Tapir IR in
Section~\ref{Sec:Implementation}. In Section~\ref{Sec:Evaluation} we describe the 
evaluation setup, including what implementations we compare to and how. We discuss
the results of evaluation in Section~\ref{Sec:Results}, including possible
explanations for discrepancies. Finally, we discuss how the many ways the work can
be extended in Section~\ref{Sec:Future}, and conclude in Section~\ref{Sec:Conclusion}.

@@

There are two classes of input sizes for the benchmarks. Some, like Floorplan,
require input files. For each of these, we chose the largest provided. For the
benchmarks with a parameter to adjust the size of the input, we attempted to 
choose a parameter size so that the fastest implementation was on the order of
seconds. The exact inputs used are:

\begin{itemize}
\item \texttt{Alignment: -f prot.100.aa} 
\item \texttt{FFT: -n 335544320}
\item \texttt{Fibonacci: -n 48}
\item \texttt{Floorplan: -f input.20}
\item \texttt{Health: -f large.input}
\item \texttt{NQueens: -n 15}
\item \texttt{Sort -n 335544320}
\item \texttt{SparseLU: -n 100 -m 100}
\item \texttt{Strassen: -n 8192}
\end{itemize}

For other implementations, we compare to GCC 7.1, Clang 4.0.1, and Intel
17.0.0. The machine used is a two socket Intel Xeon E5-2693 v3 machine with
132GB of memory, running Linux 4.11.4. All benchmarks were run using all 64
hyper-threads. 32 thread tests were done as well and showed qualitatively
similar results. It's worth noting that each of the other implementations comes
with it's own runtime. In this sense, performance is a function of both the
compiler and any optimizations it's able to perform, and the runtime.
Understanding the interaction of these two parts is non-trivial, as we will see
when discussing results.

@@

The second change in program behaviour due to our incomplete implementation was
caused by the lack \texttt{critical} pragma and \texttt{atomic} pragmas. This
issue showed up in the \texttt{Floorplan} benchmark. The \texttt{critical} pragma
ensures that only one thread can executed the referred statement at a time. 
This should be fixable in the implementation by adding a simple code-localized
synchronization generation in the IR. Similarly, the \texttt{atomic} pragma 
can be addressed by retaining a pointer to the shared stack variable, much like
existing OpenMP implementations. It is worth noting that this behaviour was
non-deterministic, and that roughly half of the time the Floorplan still
returned correct results. As we will see, this makes for an interesting
trade-off given the performance increase witnessed.

In our evaluation, we caught three kinds of errors:

\begin{itemize}
\item Segmentation faults
\item Timeouts (10 Minutes)
\item Non-determinism in result correctness
\end{itemize}

We set a timeout of 10 minutes, as at least one implementation was always
finishing within ~10 seconds, anything running more than 60 times slower
becomes irrelevant. Due to time constraints, correctness checks were not
performed on every run, so it is possible some non-determinism was missed.

@@

The failure modes enumerated in Section~\ref{Sec:Evaluation} are denoted by 
asterisks in the implementation labels on the X axis. One asterisk refers to
some incorrect answers. Two asterisks refers to time-out. Three asterisks
refers to segmentation faults.

First, we can see that there is one failure for our implementation, on
\texttt{Floorplan}. Note that the result is non-zero. In fact, while ICC and
Clang finished in roughly 1-2 seconds, the our Tapir implementation was
finishing in \emph{0.02} seconds, and computing the correct result roughly half
the time. This raises interesting questions on the cost/benefit relation of the
OpenMP behaviour that our implementation is missing.

Other failures include ICC failing with segmentation faults on \texttt{FFT} and
\texttt{Fibonacci}. While we haven't carefully diagnosed these failures, we 
suspect that the number of tasks spawned exceeded a limit in Intel's compiler.
Note that the \texttt{FFT} failures were non-deterministic, occurring on
roughly one in three runs. Clang failed with segmentation faults on every run 
of \texttt{SparseLU}. Again, we didn't have the time to investigate these
failures, and are unsure what was causing them.

The final failures we saw were timeout failures. GCC was the only culprit for these
failures. Recall that the timeout was set at 10 minutes, so any timeout means GCC 
was running at least approximately 60 times slower than the fastest implementation. 
For example, on the \texttt{FFT} benchmark, while our Tapir implementation was running
in under 5 seconds, the GCC implementation was timing out at 600 seconds, so
was at least 100 times slower.

Putting failures aside for the moment, performance for our implementation is
quite strong, generally outperforming or matching the best of existing
implementations. For benchmarks where the overhead to work ratio is high is 
where the Tapir implementation really shines. For example, for the
\texttt{Fibonacci} benchmark, our implementation finishes significantly faster
than any of the others. While ICC failed for that sized input, we did test with
smaller inputs and found its similarly outpaced by the Tapir implementation.
\texttt{NQueens} and \texttt{FFT} show similar behaviour, to lesser degrees. In 
contrast, benchmarks that have a lower overhead to work ratio unsurprisingly 
differ less in performance. For example,

\subsection{Forensics}

@@

Our primary tool for this task will be performance counters. It's worth noting
that some information is difficult to infer from performance counters, such as
sources of contention in runtimes, reasons for memory locality issues, etc.
Still, it will give us some insight into performance bottlenecks in the
program.

Before we begin understanding why, it's worth noting again that the scope of the 
work that this paper is responsible for is only the frontend. All performance
gains due to the Cilk runtime we sadly cannot claim credit for.

Another import consideration is that we can break reasons for performance
discrepancies into a few categories:

@@

While we will try out best to distinguish between these in our forensics, it is 
difficult to do so with only performance counter information. A more thorough
investigation would require careful knowledge of each of the frontends and
runtimes. One thing we can guess at is that despite the focus of the paper,
\emph{IR optimisations are unlikely to be the cause of performance
improvements}. Our evidence for this comes from Shardl et al. \cite{}. While
some IR optimisations have been implemented, they seem to have significantly
more effect on parallel loop code , while the tasking code remains
relatively unaffected by IR optimizations \cite{}.

@@

@@

@@

>>>>

The remainder of the paper is organized as follows: In Section~\ref{Sec:background} 
we give background and motivation for the paper. Following that, we describe OpenMP
tasking in Section~\ref{Sec:OpenMP}, and Tapir in Section~\ref{Sec:Tapir}. We then 
discuss the implementation, and how we compile OpenMP task constructs to Tapir IR in
Section~\ref{Sec:Implementation}. In Section~\ref{Sec:Evaluation} we describe the 
evaluation setup, including what implementations we compare to and how. We discuss
the results of evaluation in Section~\ref{Sec:Results}, including possible
explanations for discrepancies. We discuss the implications of the results ans the
significance of this work in a large context in~\ref{Sec:Discussion}. Finally,
we cover some of the many ways the work can be extended in
Section~\ref{Sec:Future}, and conclude in Section~\ref{Sec:Conclusion}.

@@

There are two classes of input sizes for the benchmarks. Some, like Floorplan,
require input files. For each of these, we chose the largest provided. For the
benchmarks with a parameter to adjust the size of the input, we attempted to 
choose a parameter size so that the fastest implementation was on the order of
seconds. The one exception was for \texttt{Fibonacci}. Because, as discussed
further below, the Intel implementation required large stacksize, we stopped at
42. The exact inputs used are:

\begin{itemize}
\item \texttt{Alignment: -f prot.100.aa} 
\item \texttt{FFT: -n 335544320}
\item \texttt{Fibonacci: -n 42}
\item \texttt{Floorplan: -f input.20}
\item \texttt{Health: -f large.input}
\item \texttt{NQueens: -n 15}
\item \texttt{Sort -n 335544320}
\item \texttt{SparseLU: -n 100 -m 100}
\item \texttt{Strassen: -n 8192}
\end{itemize}

The Intel compiler required increasing the stacksize for \texttt{Fibonacci} and
\texttt{FFT}. The implementation seemed to use significantly more stack space
than the others, requiring setting \texttt{OMP\_STACKSIZE} to \texttt{256M} for
computing the 42nd Fibonacci number. \texttt{FFT} ran fine with an increase to
\texttt{16M}.

For other implementations, we compare to GCC 7.1, Clang 4.0.1, and Intel
17.0.0. The machine used is a two socket Intel Xeon E5-2683 v3 machine with
132GB of memory, running Linux 4.11.4. Each Xeon has 16 cores running at
2.1GHz, with 20M of L3 cache. All benchmarks were run using all 64
hyper-threads. 32 thread tests were run as a sanity check and showed
qualitatively similar results. It's worth noting that each of the other
implementations comes with it's own runtime. In this sense, performance is a
function of both the compiler and any optimizations it's able to perform, and
the runtime.  Understanding the interaction of these two parts is non-trivial,
as we will see when discussing results. Each compiler was run with the flags
\texttt{-O3 -fopenmp}. Our compiler was also run with \texttt{-ftapir} to
enable Tapir instructions.

@@

The second change in program behaviour due to our incomplete implementation was
caused by the lack \texttt{critical} pragma and \texttt{atomic} pragmas. This
issue showed up in the \texttt{Floorplan} benchmark. The \texttt{critical} pragma
ensures that only one thread can executed the referred statement at a time. 
This should be fixable in the implementation by adding a simple code-localized
synchronization generation in the IR. Similarly, the \texttt{atomic} pragma 
can be addressed by retaining a pointer to the shared stack variable, much like
existing OpenMP implementations. It is worth noting that this behaviour was
non-deterministic, and that roughly half of the time the \texttt{Floorplan}
still returned correct results. As we will see, this makes for an interesting
trade-off given the performance increase witnessed.

We set a timeout of 10 minutes, as at least one implementation was always
finishing within 10 seconds, anything running more than 60 times slower becomes
irrelevant. Due to time constraints, correctness checks were not performed on
every run, so it is possible some non-determinism was missed.

@@

As mentioned earlier on \texttt{Floorplan} our implementation was sometimes
giving incorrect results. It is difficult to tell, but the time to completion
is non-zero. In fact, while ICC and Clang finished in roughly 1-2 seconds, the
our Tapir implementation was finishing in 0.02 seconds, and computing the
correct result roughly half the time. This raises interesting questions on the
cost/benefit relation of the OpenMP behaviour that our implementation is
missing. We'll discuss this further in Section~\ref{Sec:Discussion}.

GCC was the only culprit for timeouts. Recall that the timeout was set at 10
minutes, so any timeout means GCC was running at least approximately 60 times
slower than the fastest implementation. For example, on the \texttt{FFT}
benchmark, while our Tapir implementation was running in under 5 seconds, the
GCC implementation was timing out at 600 seconds, so was at least 100 times
slower. We've left GCC results off of the graphs for the three failure cases, 
\texttt{FFT}, \texttt{Fibonacci}, and \texttt{NQueens}.

Putting failures aside, performance for our implementation is quite strong,
generally outperforming or matching the best of existing implementations. For
benchmarks where the overhead to work ratio is high is where the Tapir
implementation really shines. For example, for the \texttt{Fibonacci}
benchmark, our implementation finishes significantly faster than any of the
others. While ICC failed for that sized input, we did test with smaller inputs
and found its similarly outpaced by the Tapir implementation.  \texttt{NQueens}
and \texttt{FFT} show similar behaviour, to lesser degrees. In contrast,
benchmarks that have a lower overhead to work ratio unsurprisingly differ less
in performance. For example,

\subsection{Profiling}

@@

Our primary tool for this task will be performance counters. It's worth noting
that some information is difficult to infer from performance counters, such as
sources of contention in runtimes, reasons for memory locality issues, etc.
Still, it will give us some insight into performance bottlenecks.

Before we begin understanding why, it's worth noting again that the scope of the 
work that this paper is responsible for is only the frontend. All performance
gains due to the existing Tapir backend and Cilk runtime we sadly cannot claim
credit for.

We can break reasons for performance discrepancies into a few categories:

@@

While we would like to distinguish between these, it is difficult to do so with
only performance counter information. A more thorough investigation would
require careful knowledge of each of the frontends and runtimes. One thing we
can guess at is that despite the focus of the paper, \emph{IR optimisations are
unlikely to be the cause of performance improvements}. Our evidence for this
comes from Shardl et al. \cite{}. While some IR optimisations have been
implemented, they seem to have significantly more effect on parallel loop code,
while the tasking code remains relatively unaffected by IR optimizations
\cite{}.

There are two cases we'd like to look at using the profiler. The first is a 
case when there is a large difference in runtime between implementations. For 
this case, our hypothesis is that in these cases the slower implementations
are, for some reason, spending more time in the runtime. As mentioned above, 
it is difficult to understand exactly why an implementation is spending more 
time in the runtime without further investigation, but it is a useful 
sanity check nonetheless. For this case, we turn to the \texttt{Fib} benchmark.
Indeed, we get roughly the following breakdowns of work to overhead ratio according
to the Linux \texttt{perf} tool:

\begin{itemize}
\item \texttt{tapir: ~30\% runtime overhead, ~50\% work, ~20\% other}
\item \texttt{clang: ~85\% runtime overhead, ~5\% work, ~10\% other}
\item \texttt{icc: ~85\% runtime overhead, ~5\% work, ~10\% other}
\item \texttt{gcc: ~100\% runtime overhead, ~0\% work, ~0\% other}
\end{itemize}

This confirms our suspicions that most of the slowdown incurred is overhead in
the runtime. For \texttt{tapir}, the overhead is relatively small given the
size of the work chunks, for \texttt{clang} and \texttt{icc} it is
significantly larger, while for \texttt{gcc} it is completely overwhelming.
This does let us rule out the possibility that somehow the only the codegen for
other implementations is the culprit: it is almost certainly at least partially
the runtime implementation.

For the cases where performance is close between implementations, like
\texttt{sparselu}, we expect runtimes to be dominated by actual task work. This 
expectation is confirmed by investigating performance counters for \texttt{sparselu}:
every implementation spends roughly 90\% of it's time doing work in the task
bodies.

@@

In this section we discuss the relevance of the results and this work in general
in the context of the literature. As discussed in Section~\ref{Sec:Introduction}, 
there have been many programming models for implementing parallel programs. While
this paper does not take a stance on which one \emph{should} be used, it does
accept that OpenMP is currently a widely used model, potentially the most
widely used model that has a notion of tasks.

The goal of this work is to show that a single intermediate representation can be
used to implement multiple programming models, with good performance. Schardl et 
al. showed that Tapir works well as an IR for Cilk, and surmised that it could
be an effective IR for a large subset of OpenMP. This work has taken the first
step in that direction, showing that it can work well as an IR for OpenMP tasks.

One interesting aspect of this work is that it compiles a language extension intended
for one runtime (OpenMP runtimes) to use another (the Cilk runtime). While in many
ways this is a weakness of this work, it does open the possibility of doing this 
in a general way. Having a real IR to work with enables, in the same way that LLVM
does for serial code, the possibility of multiple input parallel languages and
multiple output parallel runtimes. While the OpenMP spec's requirement to
have access to runtime routines complicates things, for many programming models
it seems at least conceivable to map them onto multiple different runtimes. This
opens the possibility of mixing programming models at compile time by targeting
the same runtime. For example, if both OpenMP and Cilk targeted Tapir
instructions, one could have a large application where some parts were written
using Cilk, while other used OpenMP, and the result would not suffer the
incompatibility problems often seen today.

@@

One of the important questions for this work is whether or not the Tapir
instructions can be extended to cover the full OpenMP semantics. As shown
in this paper, even the full semantics of OpenMP tasks weren't covered. Runtime
calls aside, implementing the full semantics of OpenMP pragmas would be a 
significant undertaking.

As two examples, let us consider the problematic \texttt{atomic} and
\texttt{critical} pragmas from the \texttt{Floorplan} benchmark. The 
\texttt{atomic} pragma would likely not be too difficult to translate
from existing OpenMP codegen. One would have to simply keep around and then
modify the contents of the original stack pointer to the shared variable. The
\texttt{critical} section would likely be a little more difficult. Still, it
should be possible to modify existing OpenMP code to block on that section.

Still, whether or not it's \emph{possible} to implement these extra features
is a separate question from whether or not it's \emph{useful}. If implementing
extra features require using special synchronization primitives, or calls into
runtimes, or any other tool other than the Tapir instructions, then the
compiler cannot reason about code using those features, and many of the
advantages we've discussed become moot, e.g. backend compatibility, generic 
optimization, etc. Features like the OpenMP \texttt{atomic} pragma don't 
require any changes to control flow, and therefore are likely to work well 
with Tapir analyses and optimizations. On the other hand features like OpenMP's
\texttt{critical} section \emph{do} require changes to control flow that seem
challenging at best to map onto Tapir's instructions. Furthermore, even if it
is possible to implement features using the Tapir instructions, it may require 
such radical entangling of the code that even if analysis is technically
possible, it becomes infeasible.

@@

There has been significant work on internal representations for parallel
programs. These can be roughly broken into four categories.

First, compilers can attach metadata to an existing IR. For example, LLVM has a
parallel loop metadata construct \cite{}. This has the benefit of being
flexible and requiring minimal work, but historically can be lost during
optimizations, and any code movement can break the semantics. The flexibility
of this approach can also be viewed as a negative, as it can compromise
what can otherwise be a simple, well defined semantics for the IR.

A second approach is to use intrinsic functions to define parallel tasks
\cite{ares}. This, like the metadata approach, has the advantage of being
flexible and relatively easy to implement, but at the cost of being less well
defined. There has recently been proposals to standardize intrinsics extensions
for LLVM to represent parallelism, but we would caution that a large, flexible
set of intrinsics being added to an already only-partially defined
\cite{verillvm} language could make reasoning about parallel programs, even
in an IR settings, infeasible.

A third approach is to have a separate instruction set for parallelism. This is
the approach taken by projects like HPIR \cite{}, SPIRE \cite{}, and INSPIRE
\cite{}. This approach has the downside of being unable to re-use existing
optimization and analysis infrastructure of an existing IR.

Finally, the last approach, and that taken by Tapir, is to implement parallel 
instructions as an extension of an existing IR. This allows for integration
into existing analyses and optimizations. In the case of Tapir, this allows one
to leverage years of development into program analysis and optimization,
extending only where necessary.

While there is general agreement that there needs to be IR support for
parallelism in parallel programs, there isn't consensus on which of these
approaches is best. Approaches that are easier at first, such as metadata
or intrinsic approaches, are similarly easy to extend to be unwieldy. There is
a reason that it's difficult to add instructions to LLVM: any added IR
construct needs to be considered in many locations. From debugging by printing
out IR, to case analysis for different control flow constructs, having 
a proper set of instructions to denote parallelism has a lot of advantages. 
Additionally, any hope of having a formal semantics for an IR depends on a 
simple, concise set of parallelism constructs, something that metadata and
intrinsics approaches can easily avoid.
